Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-19 paper.tex
Extend citation of the paper and explain differences.
Refs #8. There are other references, will have to check them too later
on. Keep it open meanwhile.

In programming courses, students solve programming problems by implementing
their solution in a particular language; in order to do this, they
must type the code, some times compile it, and finally run it to see how or if
it works. This cycle can be repeated many times, as they are in this cycle,
studens often change their affective state according to their performence.
Data can be collected while students are working on their assignments.
All the dynamics of actually typing the code and for instance
the syntax errors they received from the compiler or interpreter, the
corrections they tried, the elapsed time of each step, the number of attempts,
compiling errors, exceptions, among others. Data mining of all this data could
bring us a better understanding of the learning process, and also a better
model to both the selection of the appropriate content and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.
% to (select appropriate content| give grades| whatever ) - JJ
% I highlited the Data Ming task for the track, and that there are other options
% for Data Analytics
In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed.
The hypothesis is that students can vary the dynamic of keystrokes according
to their affective state when programming. A correlation of the
user\'s interactions (keyboard and mouse) with an emotional state, has been
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises.
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform has been proposed \cite{ijcci17}.
The platform is designed to collect keyboard and mouse data from the students\' interaction for data analysis. The proposed method has three main components: a JavaScript library to capture keyboard and mouse dynamics from a browser-based editor, a pre-processing step whose output is a feature vector that is sent to a third component, the classification algorithm.

>>>>

In programming courses, students solve programming problems by implementing
their solution in a particular language; in order to do this, they
must type the code, some times compile it, and finally run it to see how or if
it works. This cycle can be repeated many times, as they are in this cycle,
studens often change their affective state according to their performence.
Data can be collected while students are working on their assignments.
All the dynamics of actually typing the code and for instance
the syntax errors they received from the compiler or interpreter, the
corrections they tried, the elapsed time of each step, the number of attempts,
compiling errors, exceptions, among others. Data mining of all this data could
bring us a better understanding of the learning process, and also a better
model to both the selection of the appropriate content and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.

In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed.
The hypothesis is that students can vary the dynamic of keystrokes according
to their affective state when programming. A correlation of the
user\'s interactions (keyboard and mouse) with an emotional state, has been
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises.
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform was initially proposed by us in
\cite{ijcci17}. In this paper we complete the background and
motivation for this research, and update the state of the art, as well
as justify better the methdology used and the way data has been
collected. The platform we presented in that paper originally was designed to collect keyboard and mouse data
from the students\' interaction for data analysis. The proposed method
has three main components: a JavaScript library to capture keyboard
and mouse dynamics from a browser-based editor, a pre-processing step
whose output is a feature vector that is sent to a third component,
the classification algorithm.
